---
title: "ENVS 193DS Final"
author: "Kathryn Kolb"
date: "06/11/2025"
format: html
---

Github repo: https://github.com/kathrynk04/ENVS-193DS_spring-2025_final.git

```{r message=FALSE}
library(tidyverse) # loading packages
library(janitor)
library(here)
library(ggeffects)
library(MuMIn)
library(DHARMa)

sst <- read_csv(
  here("data", "SST_update2023.csv")) # reading in the data

nest_boxes <- read_csv(here("data", "occdist.csv")) # reading in the data
```

## Problem 1: Research writing

# a. Transparent statistical methods

In part 1, they used Pearson's r. In part 2, they used a one-way ANOVA.

# b. More information needed

Another test that should be included is Tukey's HSD in order to determine which source group means are different from each other. A piece of additional information that should be included is the effect size, which is the eta squared, to determine how large of an effect the source had on average nitrogen load.

# c. Suggestions for rewriting

We found a relationship (Pearson's r = correlation coefficient) between distance from headwater and annual total nitrogen load, where distance from headwater significantly predicted annual total nitrogen load (F(degrees of freedom, 1) = F-statistic, p = 0.03, $\alpha$ = significance level). We found a difference between sources in average nitrogen load (one-way ANOvA, F(4, within groups df) = F-statistic, p = 0.02, $\alpha$ = significance level).

## Problem 2: Data visualization

# a. Cleaning and summarizing

```{r message=FALSE}
sst_clean <- sst |> # storing the data into a new cleaned data frame
  clean_names() |> # cleaning the names
  mutate(year = year(date), # extracting the year from the date column
         month = month(date)) |> # extracting the month from the date column
  filter(year > 2017) |> # filtering to only include years since 2018
  mutate(month = case_match(as.character(month), # changing the month numbers to abbreviations
                            "1" ~ "Jan",
                            "2" ~ "Feb",
                            "3" ~ "Mar",
                            "4" ~ "Apr",
                            "5" ~ "May",
                            "6" ~ "Jun",
                            "7" ~ "Jul",
                            "8" ~ "Aug",
                            "9" ~ "Sep",
                            "10" ~ "Oct",
                            "11" ~ "Nov",
                            "12" ~ "Dec"),
         month = factor(month, # making month an ordered factor and selecting the levels
                        levels = c("Jan",
                                   "Feb",
                                   "Mar",
                                   "Apr",
                                   "May",
                                   "Jun",
                                   "Jul",
                                   "Aug",
                                   "Sep",
                                   "Oct",
                                   "Nov",
                                   "Dec"),
                        ordered = TRUE),
         year = factor(as.character(year), # making year a factor and selecting the levels
                       levels = as.character(2018:2023),
                       ordered = FALSE)) |>
  group_by(year, month) |> # grouping by year and month
  summarize(mean_monthly_sst = mean(temp), # calculating the mean monthly sst for each month in each year
            .groups = "drop") |>
  mutate(mean_monthly_sst = round(mean_monthly_sst, 1)) # rounding mean monthly ssts to 1 decimal place

print(str(sst_clean)) # displaying the structure of sst_clean

print(slice_sample(sst_clean, n = 5)) # displaying 5 rows from sst_clean
```

# b. Visualize the data

```{r message=FALSE}
ggplot(data = sst_clean, # making the data frame from sst_clean
       aes(x = month, # setting month as x-axis
           y = mean_monthly_sst, # setting mean_monthly_sst as y-axis
           color = year, # coloring by year
           group = year)) + # grouping by year
  geom_point() + # adding points onto plot
  geom_line() + # adding lines onto plot
  scale_color_manual(values = c("2018" = "cadetblue1", # assigning different colors to each year
                                "2019" = "cadetblue3",
                                "2020" = "cadetblue",
                                "2021" = "cadetblue4",
                                "2022" = "darkblue",
                                "2023" = "black")) +
  labs(x = "Month", # labeling the x-axis
       y = "Mean monthly sea surface temperature (°C)") + # labeling the y-axis
  theme_bw() + # making the theme black and white
  theme(panel.grid = element_blank(), # removing panel grid
        legend.position = "inside", legend.position.inside =  c(.1, 0.75)) + # putting the legend in the top left corner inside panel
  guides(color = guide_legend("Year")) # changing the legend name
```
## Problem 3: Data analysis

```{r}
#| echo: false

nest_boxes_clean <- nest_boxes |> # making a nest_boxes_clean object
  clean_names() # cleaning the names
```


# a. Response variable

In this dataset, a 1 indicates the presence of one of the three bird species, Common Starling, Swift Parrot, or Tree Martin, in a nest, and a 0 indicates that species isn't present. However, if the 1 is under the column labeled "e", then that means the nest was empty, and if there's a 0 in one of those columns, that means the nest wasn't empty.

# b. Purpose of study

The main difference between Swift Parrots and the other two species in the study is that the nests used in the study were intended to be used by Swift Parrots, because they're critically endangered. So, this study was able to determine how nest box distance from the forest edge and year impacted which species nested, therefore helping inform management in order to place nests in places that attract the target species.

# c. Difference in "seasons"

The seasons are the summer in 2016 and 2019, and the difference between these seasons was that the nests were considered new in 2016 and established in 2019.

# d. Table of models

4 models total:

| Model number | Season | Edge Distance | Predictor list              |  
|:------------:|:------:|:-------------:|-----------------------------|  
| 0            |        |               | no predictors (null model)  |
| 1            |    X   |       X       | all predictors (full model) | 
| 2            |        |       X       | edge distance               |   
| 3            |    X   |               | season                      |  

# e. Run the models

```{r message=FALSE}
model0 <- glm(sp ~ 1, # creating null model
  data = nest_boxes_clean, 
  family = "binomial")

model1 <- glm(sp ~ edge_distance + season, # creating full model (edge distance and season)
  data = nest_boxes_clean,
  family = "binomial"
)

model2 <- glm(sp ~ edge_distance, # creating model with edge distance as predictor
  data = nest_boxes_clean,
  family = "binomial"
)

model3 <- glm(sp ~ season, # creating model with season as predictor
  data = nest_boxes_clean,
  family = "binomial"
)
```

# f. Check the diagnostics

```{r message=FALSE}
par(mfrow = c(2, 2))
plot(simulateResiduals(model0)) # checking diagnostics for model0
plot(simulateResiduals(model1)) # checking diagnostics for model1
plot(simulateResiduals(model2)) # checking diagnostics for model2
plot(simulateResiduals(model3)) # checking diagnostics for model3
```

# g. Select the best model

```{r message=FALSE}
AICc(model1, # using Akaike's Information Criterion to determine which model is best
     model2,
     model3,
     model0) |> 
  arrange(AICc) # arranging in order of lowest AICc value
```
The best model as determined by Akaike's Information Criterion (AIC) that predicts the presence of Swift Parrots includes edge distance and season.

# h. Visualize the model predictions

```{r message=FALSE}
mod_preds <- ggpredict(model1, # creating the model predictions data set and naming the mmodel
                       terms =  c("edge_distance", "season")) |> # naming the predictors`
  rename(edge_distance = x, # renaming x column as "edge_distance"
         season = group) # renaming group column as "season"

ggplot(data = nest_boxes_clean, # setting up the data frame
       aes(x = edge_distance, # putting edge distance on x-axis
           y = sp, # putting swift parrot presence on y-axis
           group = season, # grouping by season
           color = as_factor(season))) + # coloring by season and making season a factor
  labs(x = "Edge Distance", # changing the x-axis label
       y = "Swift Parrot Nest Occupancy") + # changing the y-axis label
  geom_point(size = 3, # dispaying nest_boxes_clean data points
             alpha = 0.4) +
  geom_ribbon(data = mod_preds, # displaying the 95% CI for the model predictions
                  aes(x = edge_distance, # putting edge_distance on x-axis
                      y = predicted, # putting swift parrot presence prediction on y-axis
                      ymin = conf.low, # displaying the upper and lower bounds of the 95% CI
                      ymax = conf.high),
              alpha = 0.4) +
  geom_line(data = mod_preds, # displaying a line plot for model prediction
            aes(x = edge_distance, # putting edge distance on x-axis
                y = predicted)) + # putting swift parrot presence prediction on y-axis
  scale_color_manual(values = c("2016" = "pink", # coloring 2016 data pink
                                "2019" = "green")) + # coloring 2019 data green
  theme(panel.grid = element_blank()) + # getting rid of panel grid
  guides(color = guide_legend("Season")) # renaming the legend "Season"
```

# i. Write a caption for your figure

Figure 1: Model predictions and 95% confidence interval for probability of swift parrot nest occupancy according to distance from forest edge and season. Probability of swift parrot nest occupancy decreases with distance from forest edge, and probability is higher in 2016, when the nest was new, than in 2019, when the nest was established. 2016 is colored in pink and 2019 is colored in green. 95% confidence interval for both season predictions is represented by ribbons, with the 2016 ribbon colored pink and the 2019 ribbon colored green.

Data citation: Stojanovic D, Owens G, Young CM, Alves F, Heinsohn R. 2020. Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird. Restoration Ecology. 29(3). doi:https://doi.org/10.1111/rec.13319.

# j. Calculate model predictions

```{r message=FALSE}
ggpredict(model1, # predicting swift parrot nest occupancy from model1
          terms = c("edge_distance [0]", "season")) # setting edge distance as 0 and creating a prediction for each season
ggpredict(model1, # predicting swift parrot nest occupancy from model1
          terms = c("edge_distance [900]", "season")) # setting edge distance as 100 and creating a prediction for each season
```

# k. Interpret your results

The predicted probability of swift parrot nest occupancy at the forest edge (0m) is 48% for 2016 and 30% for 2019. The predicted probability of swift parrot nest occupancy farther from the forest edge (900m) is 12% for 2016 and 6% for 2019. The negative slope of the model predictions in figure 1 indicate that distance from forest edge increases, probability of swift parrot nest occupancy decreases. The relationship between distance from forest edge and probability of swift parrot nest box occupancy can be explained by the fact that one of their nest competitors, tree martins, are a native species that often takes up nest boxes that are intended for swift parrots. But, tree martins prefer to occupy nests closer to forest edges, and this competition is what drives swift parrots to select nests further from the forest edge.

## Problem 4: Affective and exploratory eisualizations

# a. Comparing visualizations

My visualizations are different from each other in the way that I've represented my data because my exploratory visualization compares in the form of boxplots the average amount of time I spent studying when I consumed coffee either while studying or not at all, while my affective visualization shows how much time I spent studying each day and how much caffeine I consumed. In my exploratory visualization, you can't tell how much caffeine I consumed each day that I studied, whereas you can tell by the size of the coffee stain in my affective visualization. Additionally, you can't tell when I consumed caffeine relative to studying in my affective visualization, but you can based on the x-axis label under each boxplot in my exploratory visualization.

My visualizations are all similar because they show how much time I spent studying each day, but that's the only thing they all have in common.

In my exploratory visualization, you can see that I spent more average time studying whenever I consumed caffeine, but it's much harder, or impossible, to see this distinction in my affective visualization. This is due to the fact that my affective visualization doesn't display the average time spent studying when I consumed caffeine and when I didn't, so it's much harder to compare based on the size of the papers and coffee stains in my affective visualization alone.

I got feedback from one of my peers to darken the shading of the coffee stains in my affective visualization, which I did end up doing and only hadn't done yet because my draft was traced in pen instead of pencil and I didn't want to cover the pencil lines, and she also suggested I include an explanation for the size of the pages, which I was able to explain in my artist statement. My instructor suggested defining the proportion of caffeine, which I was also able to define as the proportion of caffeine consumed each day in my artist statement, and that I draw zig-zags instead of straight lines on my paper. However, I wasn't able to do this because on some of the sample days I studied for over 300 or 400 minutes, so it would've taken me a really long time to measure out every centimeter of the zig-zags until I reached over 300 centimeters for multiple days, although I do agree it would've looked cool.